---
title: "Gene Ontology enrichment — ARM mutants"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
library(ggplot2)
library(GO.db)
library(org.At.tair.db)
library(clusterProfiler)
library(AnnotationDbi)
library(enrichplot)

# Install if necessary:
# BiocManager::install("clusterProfiler")
# BiocManager::install("enrichplot")
# BiocManager::install("org.At.tair.db")
# BiocManager::install("GO.db")
# BiocManager::install("AnnotationDbi")
```

## Introduction

Gene Ontology (GO) enrichment analysis was performed using clusterProfiler to identify
overrepresented biological processes (BP), molecular functions (MF), and cellular components
(CC) in the sets of differentially expressed genes (DEGs) from each comparison.

The gene universe for each comparison is defined as all genes present in the full DESeq2
results table for that comparison. DEGs are the upregulated and downregulated subsets
(padj < 0.05 and |log2FC| > 1), and GO enrichment is calculated separately for each
direction in each comparison.

---

## Setup — paths

Results from DESeq2_analysis.Rmd are read from a single result directory (one combined model,
four comparisons). GO output is written to `data/go_results/`.

```{r paths}
in_dir  <- "data/results"     # DESeq2 results (all comparisons)
out_dir <- "data/go_results"  # GO enrichment output
```

```{r setup-dirs}
dir.create(file.path(out_dir, "GO_plots"), recursive = TRUE, showWarnings = FALSE)
result_dir_pngs <- file.path(out_dir, "GO_plots")
```

---

## Functions

```{r go-function}
# Runs clusterProfiler enrichGO for a given set of target and universe TAIR IDs.
# Writes results table to out_dir and returns the enrichResult object.
calcGO <- function(universeIDs, targetIDs, outGO, onto) {

  ego <- enrichGO(
    gene          = targetIDs,
    universe      = universeIDs,
    OrgDb         = org.At.tair.db,
    keyType       = "TAIR",
    ont           = onto,
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.01
  )

  if (is.null(ego) || nrow(ego) == 0) {
    message("No significant GO terms found: ", outGO)
    return(NULL)
  }

  write.table(as.data.frame(ego), file = outGO, sep = "\t",
              col.names = TRUE, row.names = FALSE, quote = FALSE)
  return(ego)
}
```

```{r plot-function}
# Saves a barplot of the top 20 GO terms to a PNG file.
save_GO_plot <- function(ego, file_name, compare_name, suffix, onto) {

  if (is.null(ego) || nrow(ego) == 0) return(NULL)

  tryCatch({
    p <- barplot(ego, showCategory = min(20, nrow(ego))) +
         ggplot2::ggtitle(paste("GO enrichment —", suffix, "genes in", compare_name))
    ggplot2::ggsave(file_name, plot = p, width = 12, height = 8, dpi = 250)
    cat("Saved:", file_name, "\n")
  }, error = function(e) {
    cat("Error plotting", compare_name, suffix, onto, ":", e$message, "\n")
  })
}
```

```{r process-function}
# Processes GO enrichment for all three ontologies for one comparison.
process_comparison <- function(my_name, in_dir, onto) {

  df_down <- read.csv(file.path(in_dir, paste0(my_name, "_downregulated_DESeq2.csv")))
  df_up   <- read.csv(file.path(in_dir, paste0(my_name, "_upregulated_DESeq2.csv")))
  df_all  <- read.csv(file.path(in_dir, paste0(my_name, "_all_DESeq2.csv")))

  # Build universe: all genes present in the full results table for this comparison.
  all_ids     <- as.character(unique(df_all$TAIR))
  all_ids     <- all_ids[!is.na(all_ids) & all_ids != ""]
  universeIDs <- all_ids[all_ids %in% keys(org.At.tair.db, keytype = "TAIR")]

  # Filter target IDs the same way
  downIDs <- as.character(unique(df_down$TAIR))
  downIDs <- downIDs[!is.na(downIDs) & downIDs != ""]
  downIDs <- downIDs[downIDs %in% keys(org.At.tair.db, keytype = "TAIR")]

  upIDs <- as.character(unique(df_up$TAIR))
  upIDs <- upIDs[!is.na(upIDs) & upIDs != ""]
  upIDs <- upIDs[upIDs %in% keys(org.At.tair.db, keytype = "TAIR")]

  cat("\n===", my_name, onto, "===\n")
  cat("Universe:", length(universeIDs), "| Up:", length(upIDs), "| Down:", length(downIDs), "\n")

  if (length(upIDs) > 0) {
    ego_up <- calcGO(
      universeIDs = universeIDs,
      targetIDs   = upIDs,
      outGO       = file.path(out_dir, paste0(my_name, "_upregulated_GO_", onto, ".txt")),
      onto        = onto
    )
    save_GO_plot(ego_up,
      file.path(result_dir_pngs, paste0(my_name, "_GO_", onto, "_upregulated.png")),
      my_name, "upregulated", onto)
  }

  if (length(downIDs) > 0) {
    ego_down <- calcGO(
      universeIDs = universeIDs,
      targetIDs   = downIDs,
      outGO       = file.path(out_dir, paste0(my_name, "_downregulated_GO_", onto, ".txt")),
      onto        = onto
    )
    save_GO_plot(ego_down,
      file.path(result_dir_pngs, paste0(my_name, "_GO_", onto, "_downregulated.png")),
      my_name, "downregulated", onto)
  }
}
```

---

## Run GO enrichment for all comparisons

```{r go-all}
onto_list <- c("BP", "CC", "MF")

comparison_names <- sub("_upregulated_DESeq2\\.csv$", "",
                      list.files(in_dir, pattern = "_upregulated_DESeq2\\.csv$"))

for (onto in onto_list) {
  sapply(comparison_names, process_comparison, in_dir = in_dir, onto = onto)
}
```

## Session info

```{r session-info}
sessionInfo()
```
